# Formatting, Describing, and Visualizing Data

This lab discusses the basics of formatting, describing, and visualizing data. The following packages are required for this lab: 

1. car
2. reshape2
3. descr
4. psych
5. tidyverse
6. skimr

__Reminder:__ Lab One introduced how to install packages via the `install.packages()` function. 

__Note:__ These packages should already be installed from the class initialization script.

The installed packages require loading at the beginning of a R session. Remember that this can be done via the `library()` command.

## Factoring

The previous lab introduced several types of objects in R. For the purpose of this lab, objects are classified into two broad groups: factors and numerics.

Factors are nominal data that a label is applied to.
 - e.g., race, gender, party identification, etc. 

Numerics are data consisting of numbers, which are ranked (ordinal and interval). 

When data is read into R (e.g., by importing a .csv file), R automatically classifies the data by type. When data is recognized as non-numeric, R will classify it as a factor.

In the class dataset, the variable `f.party.2` is a variable consisting of factors that identifies the party affilation of the individuals who answered a survey. The `table()` function describes the variable by category.

```{r 2_factor1, echo = TRUE}
table(ds$f.party.2)
```

Attempting to take mean of the `f.party.2` variable inevitably fails.

```{r 2_mean, echo = TRUE}
mean(ds$f.party.2)
```

Party affiliation is nominal data that cannot be described via mean. Consequently, R will error when attempting to calculate the mean or median for factor variables. 

The `str()` function describes the structure of the `f.party.2` variable:

```{r 2_structure, echo = TRUE}
str(ds$f.party.2)
```

The `f.party.2` variable is a factor with three levels: "Dem", "Ind", and "Rep."

With many data sets, data are initially coded in numbers and factored afterwards. For example, the `party` variable in the `ds` data set is numeric:

```{r 2_numeric, echo = TRUE}
table(ds$party)
```

Without a codebook to decipher the numeric values, statisticians are unable to explain what 1, 2, 3, or 4 represent within the `ds` data set. Factoring data remedies this issue. Factoring data in R serves two broad purposes:

1. Applies labels to data
2. Tells R to treat the data as categorical/nominal

At a very basic level, a variable can be factored without applying labels. At minimum, R will treat the data as categorical. This method is sufficient when a variable requires quick factoring. The basic syntax is to use the `factor()` function. 
__Note:__ Best practice for factoring an existing variable is to create an additional variable within the same data set. Best practice then suggests to append the new variable with `f.` to indicate the variable is factored.

```{r 2_fac1, echo = TRUE}
ds$f.gender <- factor(ds$gender)
```

__Reminder:__ The `$` sign shows R which data set to draw the variable from, or to tell R what data set to assign the new factored variable to. The new factored variable can be described via the `table()` function.

```{r 2_fac2, echo = TRUE}
table(ds$f.gender)
```

While this factored variable is split into two categories, it is not apparent which numbers represent male and female. Labels should be assigned to the numbers to clarify the relationship between the numbers and their meaning. When factoring a variable, R requires the number of levels within the variable and the corresponding labels. In the `f.gender` variable there are two levels, 0 and 1, that require women and men labels.

```{r 2_fac3, echo = TRUE}
ds$f.gender <- factor(ds$gender, levels = c(0, 1), 
                      labels = c("Women", "Men"))
table(ds$f.gender)
```

The vector function, `c()`, tells R the levels and labels of the variable.

The `f.party` variable should also be factored, where 1 = Dem, 2 = Rep, 3 = Ind, and 4 = Other, and a table of the variable should be created.

```{r 2_fac4, echo = TRUE}
ds$f.party <- factor(ds$party, levels = c(1, 2, 3, 4), 
                     labels = c("Dem", "Rep", "Ind", "Other"))
table(ds$f.party)
```

The structure of the `f.party` variable describes a factor with four levels.

```{r 2_fac5, echo = TRUE}
str(ds$f.party)
```

There are other types of data conversions as well. In most cases, the basic syntax is the same as the `factor()` function, except the function names are `numeric()` or `integer()`.
__Note:__ For most purposes, numeric and integer are the same.

### Coerce Factoring

Sometimes the typical commands do not work and variables must be coerced. When a variable is coerced, R is instructed to treat the variable as if it were a different object type. The coercive functions are:

1. `as.factor()`
2. `as.numeric()`
3. `as.integer()`

__Try it!__ Convert a factor variable into a numeric variable using the `numeric()` function.

```{r 2_num, echo = TRUE}

```

This fails, so the coerce function should be employed.

```{r 2_num2, echo = TRUE}
ds$n.party <- as.numeric(ds$f.party)
```

Examine the new variable with the `table()` and `str()` functions.

```{r 2_num3, echo = TRUE}
table(ds$n.party)
str(ds$n.party)
```

## Recoding

In R, the `recode()` function recodes values within a variable. There many reasons to recode a variable, including:

1. To correct or change incorrect data.
2. To restructure data, making it easier for calculations.
3. To emphasize some intended effect.

For example, a recode is necessary to look at age groups instead of exact ages. If everyone in the survey reported their exact age (e.g., 54, 23, etc.) and age groups are necessary (e.g., 18-25, 26-35, etc.), the `recode()` function is useful.

To perform a recode, follow this basic syntax: `recoded.variable <- recode(old.variable, "recode commands")`

Performing a recode is best demonstrated via example. The best practice for recoding variables suggest creating a new variable with a `r.` prefix. 

Within the class data set is an `ideol` variable describing ideology. Currently, the ideology variable goes from 1 to 7, with 1 being very liberal and 7 being very conservative. The ideology variable can be recoded from seven levels to three, liberal, moderate, and conservative, via the `recode()` function. Values from 1 to 2 will recode as 1, 3 to 5 will recode as 2, and 6 to 7 will recode as 3. This `recode()` function is provided by the `car` package. 

```{r 2_recode1, echo = TRUE}
ds$r.ideol <- car::recode(ds$ideol, "1:2 = 1; 3:5 = 2; 6:7 = 3; 
                          else = NA; NA = NA")
```

The `table()` function describes the result of recoding the `ideol` variable.

```{r 2_recode2, echo = TRUE}
table(ds$r.ideol, useNA = "always")
```

__Note:__ `else = NA; NA = NA` is included at the end of the recode function. This instructs R to regard any other responses, whether missing data or data that for some reason is outside the original range, as NA, and to treat all existing NAs as NAs. Sometimes `-99 = -99` or `-99 = NA` requires inclusion in the function as well. 

__Note:__ In the recode function, all the recode arguments are required __inside one set of quotation marks__. Each argument is separated with a semicolon. R will generate an error message if commas are used. 

Using colons to define ranges will save time. In the `recode()` function, using `3:5 = 1` instructs R to recode all values between 3 and 5 to a 2. This is the preferred method opposed to typing `3 = 2; 4 = 2; 5 = 2`.

There is no standard method to categorize or recode data. Let the research question, model design, and data determine the best approach to recoding. For example, in surveys that ask individuals to support something on a scale of 1 to 4, with 1 as very supportive and 4 as least supportive, perhaps recoding the higher value to indicate greater support is appropriate.

Now let's look at the race variable:

```{r 2_race, echo = TRUE}
table(ds$race)
```

In the `ds` data set, the `race` variable consists of codes: 1 for Caucasian, 2 for African-American, and 3 through 7 for a variety of other races (Native American, Asian, Pacific Islander, 2+ races, and Other).

__Try it!__ Recode this variable to go from 7 levels to 3, where 1 is still Caucasian, 2 is African American, and 3 includes all others. 

```{r 2_race2, echo = TRUE}
ds$r.race <- car::recode(ds$race, "1 = 1; 2 = 2; 3:7 = 3; 
                         else = NA; NA = NA")
table(ds$r.race)
```

### Factoring and Recoding

Factoring a variable is generally easier subsequent to recoding. Given the `race` variable now consists of 3 levels instead of 7, factoring will add meaningful words in place of the values. The values 1, 2, and 3 can be factored as White, African-American, and Other, respectively.

```{r 2_race3, echo = TRUE}
ds$f.race.2 <- factor(ds$r.race, levels = c(1, 2, 3), 
                      labels = c("White", "African-American", "Other"))
table(ds$f.race.2)
```

The same can be done with the `ideol` variable. The values 1, 2, and 3 can be factored as Liberal, Moderate, and Conservative, respectively.

```{r 2_ideol, echo = TRUE}
ds$f.ideol <- factor(ds$r.ideol, levels = c(1, 2, 3), 
                     labels = c("Liberal", "Moderate", "Conservative"))
table(ds$f.ideol)
```

### Creating a Dummy Variable

Factoring and recoding permit the creation of dummy variables. A dummy variable is a binary indicator (0 or 1) of some category, to test for an effect from a particular category. Dummy variables are prominent in political science, so it is imperative to understand how to create and use them.

A dummy variable will be created with the recoded `race` variable to indicate whether the respondent identified as African-American (1) or not (0). Recall that in the `r.race` variable, African-American is coded as 2.

```{r 2_dummy1, echo = TRUE}
ds$r.AfAm <- car::recode(ds$r.race, "2 = 1; else = 0; NA = NA")
table(ds$r.AfAm)
```

The newly created dummy variable will now be factored to apply meaningful labels in place of the numbers.

```{r 2_dummy2, echo = TRUE}
ds$f.AfAm <- factor(ds$r.AfAm, levels = c(0, 1), 
                    labels = c("Non African-American", "African-American"))
table(ds$f.AfAm)
```

## Part III: Building and Sorting Your Data

In R, random data can be generated easily into objects and manipulated as desired. There are a variety of methods to accomplish this, the basics of which are explored below. 

The `rnorm()` function generates `n` random values that fit a normal distribution, given a mean and standard deviation.

```{r 2_norm, echo = TRUE}
one <- rnorm(100, mean = 3)
two <- rnorm(100, mean = 7)
three <- rnorm(100, mean = 1)
```

The previous code created three objects consisting of 100 random values with different means. The three objects can be combined into a single column using the `cbind()` function. The `cbind()` function will combine the  given objects sequentially in a column in the provided order. 

```{r 2_cbind, echo = TRUE}
four <- cbind(one, two, three)
```

Alternatively, the `rbind()` function can combine the three objects into a single row. Similar to the `cbind()` function, the `rbind()` function will combine the given objects sequentially in the provided order. 

```{r 2_rbind, echo=TRUE}
five <- rbind(one, two, three)
```

### The Tidyverse

The `tidyverse` is a collection of packages and functions for exploring and visualizing data in R. Developed by Hadley Wickham, the `tidyverse` packages provide a succint and consistent method of data exploration and visualization. There are plenty of different methods of working with your data, but this class will employ the `tidyverse`, as it is considered to be intuitive and rather simple. 

Let's begin with learning how to filter your data. Filtering allows you to create a subset of your data that meets specific criteria. For example, you could filter your data to examine only men, or only Republicans, etc. To filter data, use the `filter()` verb from the `dplyr` package. The `tidyverse` functions are best optimized by using the pipe operator, `%>%`, to pipe functions together. The pipe operator takes whatever is before it and sends it on to the next function. For example:

```{r tidy, echo = TRUE}
ds.men <- ds %>%
  filter(gender == 1)
```

The code chunk above creates a subset of the data that only includes men. The syntax can be read as "First take the data set, then filter it to include men only." We also assigned it to a new object, `ds.men`.  The filter verb can include multiple specifications. You also do not have to always assign your filtered data to a new object. For example, suppose you wanted to examine women who are age 42. You can do so like this:

```{r tidy2, echo = TRUE}
ds %>%
  filter(gender == 0, age == 42)
```

Similar to `filter()`, you can use the `select()` function from the `dplyr()` package to create a new data set that includes a few variables of interest.

__Note__: `select` is a very common verb in R, and therefore is often masked by other packages. If you encounter an error when using `select()`, include `dplyr::` before calling the function, so that R knows which version of `select()` to use. Follow it up by using `na.omit()` to remove missing observations:

```{r tidy3, echo = TRUE}
ds.sub <- ds %>%
  select(gender, age, income, education, ideol, glbcc_cert) %>%
  na.omit()
```


The next verb is `arrange()`, which allows you to sort your data in ascending or descending order by a particular specification. Recall the previous code chunk where we filtered the data to include only women who are age 42. Adding another pipe operator, %>%, and `arrange()` will allow us to examine the data further. Use the `arrange()` function to sort the data by education:

```{r tidy4, echo = TRUE}
ds.sub %>%
  filter(gender == 0, age == 42) %>%
  arrange(education)
```

The `slice()` function selects a set amount of observations from the data. Use `slice()` to select the first 10 observations from the data:

```{r 1tidy5, echo = TRUE}
ds.new <- ds.sub %>%
  slice(1:10)
```

The `mutate()` verb allows you to either change an existing variable or create a new one. For example, we can create a new variable that returns income in increments of 1000s:

```{r 2tidy5, echo = TRUE}
ds.new %>%
  mutate(inc_100 = income / 1000)
```

Suppose you wanted to examine summary statistics for the data. Perhaps you needed to know the average income of everyone in the survey. The `summarize()` verb can be used to do this, and more:

```{r tidy6, echo = TRUE}
ds.sub %>%
  summarize(mean_inc = mean(income))
```

Combining the `summarize()` verb with the `filter()` verb allows you to summarize a more specific set of observations. To find the average income of men only, filter the data to include only men, then summarize:

```{r tidy7, echo = TRUE}
ds.sub %>%
  filter(gender == 1) %>%
  summarize(mean_inc = mean(income))
```


Summarize can also be used to return mulitple values of interest:

```{r tidy8, echo = TRUE}
ds.sub %>%
  filter(gender == 1) %>%
  summarize(mean_inc = mean(income), med_inc = median(income), sd_inc = sd(income))
```


The code chunk above returns the mean, median, and standard deviation of income for men in the data set. 

Filtering for multiple categories can be tedious. Fortunately, there is a tidyverse function that will return desired information for each group in a variable. For example, we can get the mean, median, and standard deviation of the income variable for men and women by using the `group_by()` verb and indicating "gender":

```{r tidy9, echo = TRUE}
ds.sub %>%
  group_by(gender) %>%
  summarize(mean_inc = mean(income), med_inc = median(income), sd_inc = sd(income))
```

The `group_by` verb can be used to find lots of information. Perhaps you wanted to find the average level of belief that humans cause cliamte change by ideology level:

```{r tidy10, echo = TRUE}
ds.sub %>%
  group_by(ideol) %>%
  summarize(mean_glbcc_cert = mean(glbcc_cert))
```


The ideology variable runs from 1 to 7, with 1 being very liberal and 7 being very conservative. It is clear that liberals, on average, are more certain that humans cause climate change. The `group_by()` verb can also be used to look at multiple groups at once. We can create the same table as above, but this time break it down by ideology and gender:

```{r tidy11, echo = TRUE}
ds.sub %>%
  group_by(gender, ideol) %>%
  summarize(mean_glbcc_cert = mean(glbcc_cert))
```

### Other Methods of Exploring Your Data

There are many functions in the tidyverse that can be used for data exploration. By loading the `tidyverse` package at the beginning of the lab, most of these functions should be readily available. The `glimpse()` verb returns an overall breakdown of what is contained in your data set:

```{r tidy12, echo = TRUE}
ds.sub %>%
  glimpse()
```

In the previous section we used `summarize()` to get breakdowns of ceratin variables. If you wanted to get diagnostics on all the variables in your data, use the `skim()` function from the `skimr` package:

```{r tidy13, echo = TRUE}
skim_with(numeric = list(hist = NULL))
ds.sub %>%
  skim()
```

This provides loads of useful incormation. As demonstrated in teh previous section, you can combine different tidyverse functions together in order to maximize efficiency. For example, combining `group_by()` and `skim()` provide breakdowns of all the variables by gender:

```{r tidy14, echo = TRUE}
skim_with(numeric = list(hist = NULL))
ds.sub %>%
  group_by(gender) %>%
  skim()
```

There are a variety of operators that can be used when working with your data. These Boolean operators assist with building data subsets in R:

1. $<$ less than
2. $<=$ less than or equal to
3. $>$ greater than
4. $>=$ greater than or equal to
5. $==$ exactly equal to
6. $!=$ not equal to 
7. $!$ not (example: !x - pronounced "not x" )
8. $|$ or (example: x | y - pronounced "x OR y") 
9. & and (example: x & y - pronounced "x AND y") 

## Working with Nominal Data

Often times data is nominal. This is data that does not necessarily have a numeric value, but rather is categorized by a word or label (e.g., race, political party, etc.). The factored `party` variable is an example.

```{r 2_nom1, echo = TRUE}
table(ds$f.party)
```

If analyzing the data, the "Other" category does not explain much. In this case, recoding the Other responses as NA will exclude Other from summaries.

```{r 2_nom2, echo = TRUE}
ds$f.party.2 <- car::recode(ds$f.party, "'Dem' = 'Dem'; 'Rep' = 'Rep'; 'Ind' = 'Ind';
                            'Other' = NA; else = NA; NA = NA")
table(ds$f.party.2)
```

__Note:__ When recoding a factored variable, label names must be in in apostrophe marks __within the quotation marks__.

### Finding the Mode

When working with nominal data, there are some inapplicable statistics. For example, the mean for the factored political party variable does not exist. However, finding the mode for nominal data is useful. Recall from mathematics that the mode is the value, or in this case the label, that occurs the most often. 

A simple way to find the mode is to use the `count` verb from the `tidyverse`. Use `count` to find the mode of the factored political party variable:
```{r 2_mode, echo = TRUE}
ds %>%
  count(f.party.2)
```

Including `sort = TRUE` will tell R to sort the values in descending order, putting the modal value first:

```{r 2_mode2, echo=TRUE}
ds %>%
  count(f.party.2, sort = TRUE)
```

To get the percentage breakdown of the political parties, use the `count` verb along with the `mutate` verb. To get the percent of each category, you would take the n size of that category and divide it by the total n size!

```{r 2_mode3, echo=TRUE}
ds %>% 
  count(f.party.2) %>%
  mutate(percent = n / sum(n))
```

Perhaps you've noticed that the inclusion of NAs complicates the percentage numbers. NAs can be dropped by using the `drop_na()` function and specifying the variable of interest. Do that, and then construct the same table as above:

```{r 2_mode4, echo=TRUE}
ds %>%
  drop_na(f.party.2) %>%
  count(f.party.2) %>%
  mutate(percent = n / sum(n))
```

### Visualizing Nominal Data

R has nearly countless ways to visualize data. `ggplot2` will be the primary visualization method for these labs. There are many different visualization packages and sets of packages, but `ggplot2` provides a consistent set of visualization tools that shares a common language and syntax. A great introcution to ggplot2 can be found here:
https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

To make a visualization with `ggplot2`, first call it with the `ggplot()` function, then specify what type of visualization you want to make. The steps are:

1. Call ggplot
2. Identify the dataset (`data`)
3. Specify the aesthetic mapping (`aes`)
4. Choose what type of visualization by using the `geom_` function.

As a note, making a visualization with `ggplot2` requires multiple functions. Make sure to include a `+` sign after each function. 

A good place to start is with a bar plot. To do this, use `geom_bar()`. 

```{r 2_bar1, echo = TRUE, fig.width = 4, fig.height = 3, fig.align = 'center'}
 ggplot(ds, aes(f.party.2)) +
  geom_bar()
```

To create a visualization that only includes Democrats, Republicans, and Independents, use the `drop_na` function to filter out NAs, then pipe it to `ggplot2` Color can also be included by specifying `fill = f.party.2`. Note that inlcuding the `.` in place of the data set argument works when you pipe a dataset into `ggplot2`. Doing so tells R to use the data that is being piped into `ggplot2`.

```{r 2_bar1.5, echo = TRUE, fig.width = 4, fig.height = 3, fig.align = 'center'}
ds %>%
  drop_na(f.party.2) %>%
  ggplot(., aes(f.party.2, fill = f.party.2)) +
  geom_bar()
```

`ggplot2` has a default color pallete that it assigns to groups based on the order they appear in the data. However, matching the color red with Democrats and the color blue with Republicans might not make a lot of intuitive sense. By using the `scale_fill_manual()` function, specific colors can be specified.

```{r 2_bar2, echo = TRUE, fig.width = 4, fig.height = 3, fig.align = 'center'}
ds %>%
  drop_na(f.party.2) %>%
  ggplot(., aes(f.party.2, fill=f.party.2)) +
  geom_bar() +
  scale_fill_manual(values = c("blue", "green", "red"))
```

Construct one more visualization. This time, include a title by using `ggtitle()` and axis labels by using `xlab()` and `ylab()`.

```{r 2_bar3, echo = TRUE, fig.width = 4, fig.height = 3, fig.align = 'center'}
ds %>%
  drop_na(f.party.2) %>%
  ggplot(., aes(f.party.2, fill = f.party.2)) +
  geom_bar() +
  scale_fill_manual(values = c("blue", "green", "red")) +
  ggtitle("Breakdown of Political Parties") +
  xlab("Political Party") +
  ylab("Count")
```

## Working with Ordinal Data

Recall that ordinal data is data that is assigned numeric values, but on an ordered scale. An example of ordinal data is education level (e.g., some high school is higher than no high school, a high school diploma is higher than some high school, some college is higher than high school, etc). 

The `count()` function describes the `education` variable. Note that the `table()` function is a good way to look at categorical and ordinal data as well, but using `count()` allows us to see the NAs and is consistent with tidyverse syntax. 

```{r 2_ord, echo = TRUE}
ds %>%
  count(education)
```

The `education` variable contains 8 categories on an ordered scale. Factoring the `education` variable to apply labels will provide meaning to each value.

```{r 2_ord2, echo = TRUE}
ds$f.education <- factor(ds$education, levels=c(1, 2, 3, 4, 5, 6, 7, 8), 
				   labels=c("< HS", "HS/GED", "Vocational/Technical", 
				            "Some College", "2 year/Associates", "Bachelor's Degree", 
				            "Master's degree", "PhD/JD/MD"))
ds %>%
  count(f.education)
```

Including the `sort = TRUE` argument can return the modal education level.

```{r 2_ord3, echo = TRUE}
ds %>%
  count(f.education, sort = TRUE)
```

The `barplot()` function will visualize education levels.

```{r 2_ord4, echo = TRUE, fig.width = 6, fig.height = 4, fig.align = 'center'}
ggplot(ds, aes(f.education)) +
  geom_bar() 
```

Notice the very muddled labels. One way to solve this is to flip the coordinates of the plot by using `coord_flip()`:

```{r 2_ord4.5, echo = TRUE, fig.width = 6, fig.height = 4, fig.align = 'center'}
ggplot(ds, aes(f.education)) +
  geom_bar() +
  coord_flip()
```

More detail is available by adding colors. There are a few different methods of adding colors using ggplot. Include `fill = f.education` to indicate that the bars will be filled with color, and `scale_fill_manual()` to indiciate the colors:

```{r 2_ord5, echo = TRUE, fig.width = 6, fig.height = 4, fig.align = 'center'}
ggplot(ds, aes(f.education, fill = f.education)) +
  geom_bar() +
  scale_fill_manual(values = c("#fbb4ae", "#b3cde3", "#ccebc5", "#decbe4",
                               "#fed9a6", "#ffffcc", "#e5d8bd", "#d3d3d3")) +
  coord_flip()
```

__Note:__ The colors in the previous example are hexadecimal colors. Color schemes are available via the following website: http://colorbrewer2.org/

## Working with Interval Data

Interval data is similar to ordinal data, but with interval data the difference between levels is meaningful. A good example is age, in which the difference between 23 and 24 is the same as the difference between 56 and 57, and so on. 

The `psych` package provides a `describe()` function. The `describe()` function is useful to examine variables.

```{r 2_int, echo = TRUE}
describe(ds$age)
```

Find the modal value of `age`:

```{r 2_int2, echo = TRUE}
ds %>%
  count(age, sort = TRUE)
```

The modal value is 65.

A histogram is the appropriate visualization for interval data. Histograms returns values on a continuous scale as opposed to individual values. To make a histogram, follow the same basic steps as before, but this time use `geom_histogram()` instead.
`
```{r 2_hist, echo = TRUE, fig.width = 4, fig.height = 3, fig.align = 'center'}
ggplot(ds, aes(age)) +
  geom_histogram() 
```

To look at the density distribution, use `geom_density()` 

```{r 2_dens, echo = TRUE, fig.width = 4, fig.height = 3, fig.align = 'center'}
ggplot(ds, aes(age)) +
  geom_density()
```

To adjust the bandwidth, use `adjust =`  to specify a value.

```{r 2_dens2, echo = TRUE, fig.width = 4, fig.height = 3, fig.align = 'center'}
ggplot(ds, aes(age)) +
  geom_density(adjust = 3)
```

Last, a box plot can be generated via the `boxplot()` function. Technically a box plot in `ggplot2` requires both an x and y value, so simply indiciate `""` for the x portion of the aesthetic. 

```{r 2_box, echo = TRUE, fig.width = 4, fig.height = 3, fig.align = 'center'}
ggplot(ds, aes(x = "", y = age)) +
  geom_boxplot()
```
